\pagebreak
\thispagestyle{empty}

\begin{center}
    \textbf{\Large Preface}
\end{center}

This is not a traditional thesis. I did not follow a single well defined
research line, there was not a single well defined methodology, and I did not
have a single well defined goal. I don't think that this is inherently bad, and
perhaps this situation is indeed increasingly common, but yet we are asked to
write a thesis that follows a traditional structure and that tells a story that
is not necessarily the one that we have lived. Here I would like to briefly
explain the factors that lead to the diversity of topics in this thesis.

First of all, my own personality: curious, open-minded... sorry, this is just
an insider joke, look at the IFISC lema! Now seriously, my curiosity, and
perhaps my little patience, made me jump from one topic to another, from one
methodology to another, and from one goal to another. Second, less joyful,
the lack of funding: I never obtained a grant for my PhD. Every year, even if
not actively, I had to think on how to get funding for the next year, nothing
was taken for granted. This made me go into each and almost every opportunity
that appeared. But don't get me wrong, I really like working on different
topics, just that this does not help to build a traditional thesis. Here I must
say that I was not alone in this situation, my supervisor had always several
plans for possible funding sources and he always gave me the freedom to choose
on what to work on. Sadly, this is not the case for many PhD students, with or
without grants. I am conscious that I was truly lucky. Finally, the nature of
the scientific system that has been built over the last decades:
publish or perish, impact factor, h-index, etc. I do not consider that I
have made ``bad'' science, but perhaps I would have continued doing research on
some topics that I abandoned if I had not been ``worried'' about the impact of
my work.

At any rate, throughout the pages of this thesis and especially in the
Introduction, I will try to convince you that all the work I have done is
connected, that there is a common thread that connects all the topics I have
worked on. Despite I belive that this is true (I am of course not lying to
you), if we are to be honest, this takes a secondary role. I don't want to
steal more of your time, I will already do it with the rest of the thesis,
which is not characterized by its brevity.

To finish, just to comment that I have tried to write this thesis for all
audiences, not only for the experts in the field. I have tried to explain the
concepts in a simple way, and I have avoided the use of jargon as much as
possible, especially mathematical one. I have also tried to make it enjoyable
to read, with some more personal comments.

I hope you enjoy reading this thesis as much as I enjoyed when I saw it
finished (I wouldn't say ``as I enjoyed it writting it'', precisely) and that
you find it interesting and inspiring.

\vfill